---
title: "Mar Casado PCA Validation"
format: html
editor: visual
execute: 
  message: false
  warning: false
---

The aim of this document is to validate results of the ml.js package PCA method in JavaScript for an Observable course on [Data Science Workflows in JavaScript (Session 3)](https://observablehq.com/@observablehq/ds-workflows-in-js-session-3-key). This example closely follows the example from the [palmerpenguins package documentation](https://allisonhorst.github.io/palmerpenguins/articles/pca.html).

**Data source:** Marcos Gonçalves da Silva, Juliana Nascimento Silva, Helena Rodrigues Fragoso, Natalia Pirani Ghilardi-Lopes (2019).Temporal series analysis of abiotic data for a subtropical Brazilian rocky shore. Data in Brief, Volume 24. ISSN 2352-3409, <https://doi.org/10.1016/j.dib.2019.103873>.

### Attach packages

```{r}
library(corrr)
library(GGally)
library(recipes)
library(tidytext)
library(tidyverse)
library(lubridate)
```

### Get data

Read in the CSV, and add a column indicating "hot moist" and "cool dry" seasons according to authors.

```{r}
mc <- read_csv("marCasadoData.csv") |> 
  mutate(season = dplyr::case_when(lubridate::month(month) %in% c(10, 11, 12, 1, 2, 3) ~ "hot moist", TRUE ~ "cool dry"))
```

### Explore correlations:

```{r}
marCasadoCorr <-  mc %>%
  dplyr::select(-month, -season) %>%
  correlate() %>%
  rearrange()
```

### Make the PCA recipe:

```{r}
mc_recipe <-
  recipe(~., data = mc) %>% 
  update_role(month, season, new_role = "id") %>% 
  step_normalize(all_predictors()) %>%
  step_pca(all_predictors(), id = "pca") %>% 
  prep()
```

### Do PCA:

```{r}
mcPCA <- 
  mc_recipe %>% 
  tidy(id = "pca") 
```

### Create a table of loadings:

```{r}
mcPCA |> select(-id) |> pivot_wider(id_cols = "terms", names_from = "component", values_from = "value")
```

### Visualize varianced explained by PC

```{r}
mc_recipe %>% 
  tidy(id = "pca", type = "variance") %>% 
  dplyr::filter(terms == "percent variance") %>% 
  ggplot(aes(x = component, y = value)) + 
  geom_col() + 
  xlim(c(0, 10)) + 
  ylab("% Total Variance") +
  theme_minimal()
```

### Create biplot

```{r}

pca_wider <- mcPCA %>% 
  tidyr::pivot_wider(names_from = component, id_cols = terms)

arrow_style <- arrow(length = unit(.05, "inches"),
                     type = "closed")

pca_plot <-
  juice(mc_recipe) %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(alpha = 1, 
             size = 3,
             aes(color = season)) + 
  scale_color_manual(values = c("steelblue", "darkorange"))

pca_plot +
  geom_segment(data = pca_wider,
               aes(xend = 5 * PC1, yend = 5 * PC2), 
               x = 0, 
               y = 0, 
               arrow = arrow_style) + 
  geom_text(data = pca_wider,
            aes(x = 5 * PC1, y = 5 * PC2, label = terms), 
            hjust = 0, 
            vjust = 1,
            size = 3, 
            color = '#0A537D') + 
  theme_minimal()
```

\
